---
title: "Global deposition of Beryllium in ice cores"
subtitle: "Draft Report"
author: "Hoang Long Nguyen and Dmytro Perepolkin"
date: "2023-01-25"
output: 
  html_document: 
    highlight: pygments
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data description

The data file production_data.csv contain 400 observations on 8 variables. 6 of 
the variables are synthetic production rate of 10Be for the last 8000 years from 
different regions of the globe. These regional synthetic production rate are 
based on a production model that can simulate theoretical production rate at 
each latitude and height (km). We generate the production rate and combine them 
(summation) into different regions. Therefore, correlation are expected between 
neighboring regions. Beside regional correlation, there are also time 
autocorrelation in the data set.

## Variable description
age           time before 1950 [years]
Q_trop_high   10Be production rate in the troposphere at high latitude (60-90°) [atoms/cm2/s]
Q_trop_mid    10Be production rate in the troposphere at mid latitude (30-60°) [atoms/cm2/s]
Q_trop_low    10Be production rate in the troposphere at low latitude (0-30°) [atoms/cm2/s]
Q_stra_high   10Be production rate in the stratosphere at high latitude (60-90°) [atoms/cm2/s]
Q_stra_mid    10Be production rate in the stratosphere at mid latitude (30-60°) [atoms/cm2/s]
Q_stra_low    10Be production rate in the stratosphere at low latitude (0-30°) [atoms/cm2/s]
EDML          10Be flux [10^6 atoms/cm2/year] to Antarctica measured at the EDML ice core [normalized]
GRIP          10Be flux [10^6 atoms/cm2/year] to Greenland measured at the GRIP ice core [normalized]

The variables EDML and GRIP (10Be flux) were normalized to have the same mean as the sum of 10Be production rate from all the regions over the last 8000 years.
The goal is to analyse how much each region contributes to the 10Be fluxes to either Greenland or Antarctica


When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document. You can embed an R code chunk like this:

```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
